\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-06-30 tex/paper.tex
save

In this work, we propose to rely on variational optimization to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q(\theta|\psi)$ over the
parameters of $g$ and $p(\mathbf{x}|\theta)$ and minimize in alternation the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\phi$ and $\psi$.
When updating
$\phi$, unbiased estimates of $\nabla_\phi U_d$ can be obtained by
evaluating the exact and known gradient of $U_d$ over mini-batches of true and
generated data, as ordinarily done in stochastic gradient descent. When updating
$\psi$, estimates of $\nabla_\psi U_g$ can be derived with forward
simulations, as described in the previous section.
That is,
\begin{equation}
    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q(\theta|\psi), \mathbf{z} \sim p(\mathbf{z}|\theta)}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q(\theta|\psi)],
\end{equation} which we can approximate with mini-batches of
generated data
\begin{equation}
    \nabla_\psi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q(\theta_m|\psi)
\end{equation}
for $\theta_m \sim q(\theta|\psi)$ and $\mathbf{z}_m \sim p(\mathbf{z}|\theta_m)$.
For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
optimization procedure, as built on top of WGAN-GP.
Obviously, the variational relaxation could similarly be coupled with
other variants of GANs and/or of evolution strategies.

@@

The top left plot in Figure~\ref{fig:poisson} illustates the resulting proposal
distributions $q(\theta|\psi)$ for $\gamma=0$ and $\gamma=5$ after adversarial
variational optimization. We see that the distributions both arrive in the
neighborhood of $\theta^*$, with a density highly more concentrated for
$\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
distance with $\theta^*$, both models closely match the observed data, as shown
in the top right plot of  Figure~\ref{fig:poisson}. This suggests either a
relatively flat landscape around $\theta^*$ or that the observed data can in
this case also be reproduced with a smeared variation $p(\mathbf{x}|\psi)$ of
the simulator. Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that
for both $\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ quickly
fall towards $0$, which indicates  convergence towards a distribution that the
critic cannot distinguish from the true distribution.

@@

\glnote{avo = extension of gans to non-differentiable generator}
\glnote{nice for discrete data}
\glnote{next version = try on an actual particle physics simulator.}
\glnote{nes for better invariance and stability?}

\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$
is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?}

>>>>

In this work, we propose to rely on variational optimization to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q(\theta|\psi)$ over the
parameters of $g$ and $p(\mathbf{x}|\theta)$ and minimize in alternation the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\phi$ and $\psi$.
When updating
$\phi$, unbiased estimates of $\nabla_\phi U_d$ can be obtained by
evaluating the exact and known gradient of $U_d$ over mini-batches of true and
generated data, as ordinarily done in stochastic gradient descent. When updating
$\psi$, estimates of $\nabla_\psi U_g$ can be derived with forward
simulations, as described in the previous section.
That is,
\begin{equation}\label{eqn:grad-ug-approx}
    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q(\theta|\psi), \mathbf{z} \sim p(\mathbf{z}|\theta)}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q(\theta|\psi)],
\end{equation} which we can approximate with mini-batches of
generated data
\begin{equation}
    \nabla_\psi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q(\theta_m|\psi)
\end{equation}
for $\theta_m \sim q(\theta|\psi)$ and $\mathbf{z}_m \sim p(\mathbf{z}|\theta_m)$.
For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
optimization procedure, as built on top of WGAN-GP.
Obviously, the variational relaxation could similarly be coupled with
other variants of GANs and/or of evolution strategies.

@@

The top left plot in Figure~\ref{fig:poisson} illustates the resulting proposal
distributions $q(\theta|\psi)$ for $\gamma=0$ and $\gamma=5$ after adversarial
variational optimization. We see that the distributions both arrive in the
neighborhood of $\theta^*$, with a density highly more concentrated for
$\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
distance with $\theta^*$, both models closely match the observed data, as shown
in the top right plot of  Figure~\ref{fig:poisson}, with again
slighlty better results for the entropy penalized model. This suggests either a
relatively flat landscape around $\theta^*$ or that the observed data can in
this case also be reproduced with a smeared variation $p(\mathbf{x}|\psi)$ of
the simulator. Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that
for both $\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ quickly
fall towards $0$, which indicates  convergence towards a distribution that the
critic cannot distinguish from the true distribution.

@@

In this note, we develop a likelihood-free inference algorithm for fitting a
forward non-differentiable simulator to continuous or discrete observed data.
The algorithm combines adversarial training  with variational optimization to
minimize variational upper bounds  on the otherwise non-differentiable
adversarial objectives.
Effectively, the procedure results in learning an abitrarily tight proposal
distribution over the simulator parameters, such that corresponding the marginal
distribution of the generated data matches the observations. Preliminary results
with non-differentiable discrete or continuous simulators validate the proposed method.

While the obtained results are encouraging, the complete validation of the
method remains to be carried out in real conditions on a full fledged scientific
simulator -- which we plan to achieve for a next version of this work.
In terms of method, several components need to be further investigated.
First, we need to better study the interplay between the entropy penalty and the adversarial objectives.
Second, we should better understand the dynamics of the optimization
procedure, in particular when combined with momentum-based optimizers like Adam.
Third, we need to consider whether less noisy estimates of the gradients
$\nabla_\psi U_g$ can be computed, in particular by exploiting the fact in the composition $d(g(\mathbf{z};\theta); \psi)$
the derivatives of $d$ are known exactly.
